Peer Assessment 2 - Exploratory Data Analysis
========================================================
<p>
Code for peer assessment 2 - Exploratory Data Analysis
</p>
```{r echo=FALSE, message=FALSE, warning=FALSE}
library(plyr)
library(ggplot2)
library(gridExtra)
```



## Data Processing and Exploration
```{r,echo=TRUE, cache=TRUE}
# Read the rd file contents
setwd("F:/My Development/My Testing Ground/Data Science/Data Science Cert - Code/Exploratory Data Analysis/Week 3")
NEI = readRDS("summarySCC_PM25.rds")
SCC = readRDS("Source_Classification_Code.rds")
```

```{r,echo=TRUE}
# summary of NEI
str(NEI)

# summary of SCC
str(SCC)

# top 10 entries in NEI
head(NEI, 10)

# top 10 entries in SCC
head(SCC,10)

# description for 10100101
SCC[which(SCC$SCC == "10100101"),]
```


### Question 1 : Have total emissions from PM2.5 decreased in the United States from 1999 to 2008? Using the base plotting system, make a plot showing the total PM2.5 emission from all sources for each of the years 1999, 2002, 2005, and 2008.?
<p>
Let's utilize a plot depicting the total emissions, over the years, across all sources to understand the fluctuation of emission from PM2.5.
</p>
```{r,echo=TRUE, cache=TRUE}
# Group emissions by year across all sources and find the total emission for each of the years
gdf = ddply(NEI, c("year"), function(x) c(TOTAL=sum(x$Emissions)))
```

```{r, echo=TRUE}
plot(gdf$year,gdf$TOTAL, type = "o", ylim = c(0, max(gdf$TOTAL)), xlim=c(1998, 2010),
     col="red",
     xlab="Year",ylab="Total PM2.5 Emission (Tons)",
     main="Total PM2.5 Emissions (in Tons) from Year 1999 to 2008")
```

<p>
From the above graph it is clearly evident that emissions from PM2.5 has decreased from year 1998 to 2008
</p>



### Question 2 : Have total emissions from PM2.5 decreased in the Baltimore City, Maryland (fips == "24510") from 1999 to 2008? Use the base plotting system to make a plot answering this question.
```{r, echo=TRUE}

# Extract emission data for Baltimore City (fips == "24510")
gdf = NEI[which(NEI$fips == "24510"),]

# For the city of Baltimore, group emissions by year across all sources and find the total emission for each of the years
gdf = ddply(gdf, c("year"), function(x) c(TOTAL=sum(x$Emissions)))

# Plot
plot(gdf$year,gdf$TOTAL, type = "o", ylim = c(0, max(gdf$TOTAL)), xlim=c(1998, 2010),
     col="red",
     xlab="Year",ylab="Total PM2.5 Emission (Tons)",
     main="Total PM2.5 Emissions (in Tons) for Baltimore City from Year 1999 to 2008")
```

<p>
From the above graph it is clearly evident that in the City of Baltimore emission from PM2.5 decreased from year 1998 to 2008. However, the fall in the emission rate wasn't steady. As you can in the figure, the emission did decrease from 1998 to 2002, but bumped back up in 2005 before finally receding to lowest level in 2008.
</p>



### Question 3 : Of the four types of sources indicated by the type (point, nonpoint, onroad, nonroad) variable, which of these four sources have seen decreases in emissions from 1999–2008 for Baltimore City? Which have seen increases in emissions from 1999–2008? Use the ggplot2 plotting system to make a plot answer this question.
```{r, echo=TRUE}

# Extract emission data for Baltimore City (fips == "24510")
gdf = NEI[which(NEI$fips == "24510"),]

# For the city of Baltimore, group emissions by year across all sources and find the total emission for each of the years
gdf = ddply(gdf, c("type","year"), function(x) c(TOTAL=sum(x$Emissions)))

# Plot
ggplot(gdf, aes(x=year, y=TOTAL, colour=type)) + geom_line() +
    xlab("Year") +
    ylab("Total PM2.5 Emission (Tons)") +
    ggtitle("Total Emission from PM2.5 in Baltimore City \nFrom Different Sources During the Year 1998-2008") +
    scale_colour_discrete(name = "Emission \nSource Type")+ 
    geom_point(colour="red", size=4, shape=21, fill="white")

```

<p>
Sources that have seen DECREASES in emission from 1999-2008 for Baltimore City:
NONPOINT
NON-ROAD
ON-ROAD

Sources that have seen INCREASES in emission from 1999-2008 for Baltimore City:
POINT
</p>


### Question 4 : Across the United States, how have emissions from coal combustion-related sources changed from 1999–2008?
```{r, echo=TRUE}
# Find all coal sources
sccCoal = SCC[grep("Coal+", SCC$EI.Sector, perl=TRUE, value=FALSE),]

# Find emission of PM2.5 from coal sources that was recorded through out USA
gdf = NEI[which(NEI$SCC %in% sccCoal$SCC),]

# Group emissions by year produced by coal sources through out USA
gdf = ddply(gdf, c("year"), function(x) c(TOTAL=sum(x$Emissions)))

# Plot
ggplot(gdf, aes(x=year, y=TOTAL)) + geom_line(color="red") +
    xlab("Year") +
    ylab("Total PM2.5 Emission (Tons)") +
    ggtitle("Total Emission from Coal Sources\n Through Out USA During the Year 1998-2008")+ 
    geom_point(colour="red", size=4, shape=21, fill="white")


```

<p>
Across United States, emissions from Coal sources have decreased considerably through the years 1998-2008.
</p>




### Question 5 : How have emissions from motor vehicle sources changed from 1999–2008 in Baltimore City?
```{r, echo=TRUE}

# Find all motor vehicle emission sources
sccCoal = SCC[which(SCC$Data.Category=="Onroad"),]

# Find emission of PM2.5 from motor vehicle sources that was recorded through out USA
gdf = NEI[which(NEI$SCC %in% sccCoal$SCC),]

# Extract emission data for Baltimore City (fips == "24510")
gdf = gdf[which(gdf$fips == "24510"),]

# Group emissions by year produced by motor vehical sources in Baltimore City
gdf = ddply(gdf, c("year"), function(x) c(TOTAL=sum(x$Emissions)))

# Plot
ggplot(gdf, aes(x=year, y=TOTAL)) + geom_line(color="red") +
    xlab("Year") +
    ylab("Total PM2.5 Emission (Tons)") +
    ggtitle("Total Emission from Motor Vehicle Sources\n in Balitmore City During the Year 1998-2008")+ 
    geom_point(colour="red", size=4, shape=21, fill="white")


```

<p>
In Baltimore City, emissions from Motor Vehicle sources have decreased considerably through the years 1998-2008.
</p>



### Question 6 : Compare emissions from motor vehicle sources in Baltimore City with emissions from motor vehicle sources in Los Angeles County, California (fips == "06037"). Which city has seen greater changes over time in motor vehicle emissions?
```{r, echo=TRUE}

# Find all motor vehicle emission sources
sccCoal = SCC[which(SCC$Data.Category=="Onroad"),]

# Find emission of PM2.5 from motor vehicle sources that was recorded through out USA
gdf = NEI[which(NEI$SCC %in% sccCoal$SCC),]

# Extract emission data for Baltimore City (fips == "24510") and Los Angeles County (fips == "06037")
gdf = gdf[which(gdf$fips == "24510" | gdf$fips == "06037"),]

# Assign proper names for 'fips'
gdf[which(gdf$fips == "24510"),]$fips = "Baltimore City"
gdf[which(gdf$fips == "06037"),]$fips = "Los Angeles County"

# Group emissions by fips (Baltimore & Los Angeles) and year produced by motor vehical sources
gdf = ddply(gdf, c("fips","year"), function(x) c(TOTAL=sum(x$Emissions)))

# Plot
ggplot(gdf, aes(x=year, y=TOTAL, colour=fips)) + geom_line() +
    xlab("Year") +
    ylab("Total PM2.5 Emission (Tons)") +
    ggtitle("Total Emission from Motor Vehicles in Baltimore City \n vs Los Angeles County During the Year 1998-2008") +
    scale_colour_discrete(name = "US County")+ 
    geom_point(colour="red", size=4, shape=21, fill="white")
```

<p>
Motor vehicle emission has been considerably high in Los Angeles county since 1998. More over, compared to the same year (1998) in Baltimore City the emission was 10 time higher in Los Angeles county. For Baltimore City the emission rate has decreased steadly through 1998-2008, where as for Los Angeles County the rate increased tremendously, hitting the highest point in 2005. Even though the emission rate did decrease for Los Angeles after 2005, the rate as of 2008 is still higher than what it was in 1998.
</p>