US National Weather Service Storm Data: Economical and Personal Costs Analysis
========================================================

<p>
I'm copying an RPub publication from my fellow Coursera student. Here's the full article: http://rpubs.com/PedroCebrian/RepResPeer2
For sake of brevity, I'm only including the code. Read through the original article if you need further information.
</p>

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(plyr)
library(ggplot2)
library(gridExtra)
setwd("F:/My Development/My Testing Ground/Data Science/Data Science Cert - Code/Reproducible Research/Week 4")
```


## Data proccessing

<p>
The first task is to unzip and read the data from the downloaded file.
</p>

```{r echo=TRUE}
fname = "repdata-data-StormData.csv.bz2"
con = bzfile(fname)
data = read.csv(con)
unlink(con)
```

```{r echo=TRUE}
dataVars = data[ ,c("EVTYPE","FATALITIES","INJURIES","CROPDMG",
                      "CROPDMGEXP","PROPDMG","PROPDMGEXP")]

rm(data) # remove the data variable that was created above

# Assign numbers for each of the magnitudes (k, K, h, H etc).
# Note, this is the key piece that I missed. I have a nested loop
# which was very inefficient and ugly.
levels(dataVars$PROPDMGEXP)<- c(1,1,1,1,1,10,100,1000,10000,
                                 1000000,1000000,10000000,
                                 100000000,1000000000,100,100,
                                 1000,1000000,1000000)

levels(dataVars$CROPDMGEXP)<- c(1,1,1,100,1000000000,1000,1000,
                                 1000000,1000000) 

# Convert the above maginitudes to numbers
dataVars$PROPDMGEXP <- as.numeric(as.character(dataVars$PROPDMGEXP))
dataVars$CROPDMGEXP <- as.numeric(as.character(dataVars$CROPDMGEXP))



dataFinal <- dataVars[ ,c("EVTYPE","FATALITIES","INJURIES")]
dataFinal$FATALITIES.INJURIES <- with(dataVars, {FATALITIES + INJURIES})
dataFinal$CROP.DAMAGES <- with(dataVars, {CROPDMG * CROPDMGEXP})
dataFinal$PROP.DAMAGES <- with(dataVars, {PROPDMG * PROPDMGEXP})
dataFinal$DAMAGES <- with(dataVars, {PROPDMG * PROPDMGEXP + CROPDMG * CROPDMGEXP})

```



```{r echo=TRUE}
total =  tapply(dataFinal$CROP.DAMAGES, dataFinal$EVTYPE, sum)
total = sort(total, decreasing = TRUE)
total
```

